---
title: "EDS 223 HW#1"
format: html
warning: False
---

# *Assessing proximity of hazardous waste disposal sites to low income communities*


#### Author - Henry Oliver

## Purpose:
The purpose of this document is to practice geospatial data analysis by creating visually appealing maps that display an environmental injustice in Michigan's Murray County.

## Workflow:

### 1)  Load in the relevant libraries

```{r}

library(tidyverse)
library(sf)
library(here)
library(tmap)
library(stars)
library(terra)
```

### 2)  Read in geodatabase of EJScreen data at the Census Block Group level

```{r}
#| message: False
ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"), quiet = TRUE)
```

### 3)  Read in other data to be used in map display

```{r}
#| message: False
# Shapefile for cities in detroit, filtered for most populous
mi_cities <- sf::st_read(here::here("data", "tl_2017_26_place", "tl_2017_26_place.shp"), quiet = TRUE) %>% 
  st_transform(crs = st_crs(ejscreen)) %>% # Transform CRS for inset map
  filter((NAME == "Detroit" )| (NAME == "Dearborn"))

# Read in shapefile for the great lakes
great_lakes <- sf::st_read(here::here("data", "GL250515_lam", "GL250515_lam.shp"), quiet = TRUE) %>% 
  st_transform(crs = st_crs(ejscreen))
```

```{r}
#| echo: False
if (st_crs(mi_cities) == st_crs(ejscreen)) {
  print("CRS for ejscreen and mi_cities are matching")
} else {
  print("CRS DOES NOT MATCH")
}

if (st_crs(great_lakes) == st_crs(ejscreen)) {
  print("CRS for ejscreen and great_lakes are matching")
} else {
  print("CRS DOES NOT MATCH")
}
```

### 4)  Summarize geometries to get just state borders (for inset map)

```{r}
states <- ejscreen %>% 
  group_by(ST_ABBREV) %>% # Group by state
  summarise() %>% # Summarise so there's only one geometry per state
  ungroup()
```

### 5)  Further filter data

```{r}
# filter to Michigan
michigan <- ejscreen %>%
  dplyr::filter(ST_ABBREV == "MI") 

# Create a shapefile for just the counties
mi_counties <- michigan %>% 
  group_by(CNTY_NAME) %>%   # Group by counties
  summarise() %>% # Summarize geometries(default) for each county
  ungroup()

# filter to Wayne County and surrounding counties in map extent
wayne <- michigan %>%
  dplyr::filter(CNTY_NAME %in% c("Wayne County",
    "Oakland County",
    "Macomb County",
    "Washtenaw County",
    "Monroe County"))

wayne_inset <- st_union(wayne, by = 'Shape')

```

### 6)  Create bounding boxes for primary map(s) and inset map(s)

```{r}
# Set bounds for primary display maps
bounds <- st_bbox(c(xmin = -83.6, xmax = -82.63, ymin = 42.02, ymax = 42.5))

# Set bounds for inset map
state_bounds <- st_bbox(c(xmin = -89, xmax = -76 , ymin =  39, ymax = 47))


```

### 7)  Create inset map

```{r}
# Create inset map
# Set bounds to states
inset_map <- (
tm_shape(states, bbox = state_bounds) +
  tm_polygons(fill = "#fef0d9") +
  tm_text("ST_ABBREV", # Add text to include state abbreviation
          size = 0.75, # Set text size
          col = "black", # Color
          fontface = "bold", # Make it bold
          xmod = 1.35, ymod = -0.75) + # Change the position of text relative to centroid (so we can see it)
tm_shape(great_lakes) + # Add the great lakes shape
  tm_polygons(fill = "lightblue")) +
  tm_shape(wayne_inset) +
  tm_borders(col = 'red', lwd = 3)

```

### 8)  Create palette for display

```{r}
# Create palette for hazardous waste sites and % income
palette = palette = (c('#fef0d9',
                    '#fdcc8a',
                    '#fc8d59',
                    '#d7301f',
                    '#8b0000'))
```

### 9)  Create map displaying proximity to Hazardous Waste Sites

```{r}

tm_shape(wayne, bbox = bounds) + # Plot county polygons within bounds
  tm_polygons(fill = 'PTSDF',
            breaks = c(0, 1, 3, 5, 7, Inf), # Set bin breaks
            palette = palette, # Set color palette
            lty = 'blank', # Remove census borders,
            title = "Sites within 5km", # Title legend
            labels = c("0", "1-2", "3-5", "5-7", "7+")) + # Determine fill param
  tm_shape(mi_counties, bbox = bounds) + # Add county borders
  tm_borders() +
  tm_shape(great_lakes)+ # Add great lakes borders
  tm_polygons(fill = "lightblue") +
   tm_shape(mi_cities) + # Determine shape for most populous cities
  tm_dots() + # Add points for most populous cities
  tm_text(text = 'NAME', ymod = 0.5, fontface = 'bold', size = 0.75) +
  tm_title(text = "Fig 1: Proximity to Hazardous Waste Sites in Wayne County, MI", # Add title
            size = 3,
            position = c(0, 1.1)) + # Specify title position
  tm_scalebar(position = c(0, 0)) +
  tm_compass(position = c(0.65, 0.32), size = 3) +
  tm_inset(inset_map,   # Add inset map
           width = 10,
           height = 9, # Set size and position of inset map
           position = c(0.8, 0.5))
```

### 10) Create map displaying percentage of low income households per census tract

```{r}
tm_shape(wayne, bbox = bounds) + # Plot county polygons within bounds
  tm_polygons(fill = 'LOWINCPCT', # Determine fill parameter
              lty = 'blank',
              palette = palette,
              title = "% Low Income",
              labels = c("0 - 19.99", "20 - 39.99", "40 - 59.99", "60 - 79.99", "80 - 100")) + 
  tm_shape(mi_counties, bbox = bounds) + # Add county borders
  tm_borders() +
  tm_shape(great_lakes)+ # Add great lakes borders
  tm_polygons(fill = "lightblue") +
  tm_title(text = "Fig 2: Percentage of Low Income Households in Wayne County, MI", # Add title
            size = 3,
            position = c(0, 1.1)) + 
  tm_shape(mi_cities) + # Determine shape for most populous cities
  tm_dots() + # Add points for most populous cities
  tm_text(text = 'NAME', ymod = -0.5, col = 'black', fontface = 'bold', size = 0.75 ) + # Add text above most populous cities
  # Specify title position
  tm_scalebar(position = c(0, 0)) +
  tm_compass(position = c(0.65, 0.32), size = 3) +
  tm_inset(inset_map,   # Add inset map
           width = 10,
           height = 9, # Set size and position of inset map
           position = c(0.8, 0.5),
           frame = FALSE)

 
```

## Summary:

  Fig 1 above shows the concentration of hazardous waste disposal sites (HWDS) for census blocks in Wayne County, MI, where a darker shade of red on the map corresponds to a higher number of HWDSs within 5 kilometers (km) of each census block. Fig 2 shows the percentage of low income households per census block, where a darker shade of red corresponds to higher percentage of low-income households. As we see on the maps, there are high clusters of both HWDS and low-income households 5km Southeast of Dearborn, and about 5km southeast of Detroit.  Furthermore, there are significant clusters of both variables on the coast about 20km southeast of Dearborn. Clusters appear less congruent in the western portion of the county, although 20km west of Detroit there is a high concentration of low-income households that correspond with a higher density of HWDS. At the very least, these figures show that the concentrations of low income households and hazardous waste disposal are both highest in northeast Wayne County, near Detroit. Though hazardous waste site density could likely be correlated more closely with population density, the correlations between the two variables plotted above encourage more research towards my hypothesis that low-income households in Murray county are more likely to be located near HWDSs than their higher-income counterparts
  
  
  
### References:

U.S. Environmental Protection Agency (EPA), 2023. EJScreen Technical Documentation.

U.S. Census Bureau. (n.d.). TIGER/Line shapefile, 2017, Michigan current place (state‑based) [Data set] https://catalog.data.gov/dataset/tiger-line-shapefile-2017-state-michigan-current-place-state-based

U.S. National Ice Center. (n.d.). Great Lakes data support files. Retrieved October 6, 2025, from https://usicecenter.gov/Products/GreatLakesData